---
title: 'Lab 3: String processing'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(pdftools)
library(stringr)
library(lubridate)
```


In this lab we'll be learning more about string processing using `stringr`, and `pdftools`. You'll also need the packages `lubridate`, `tidyverse`, and `dplyr`.

### Problem 1: Birthdays

For Problem 1 we are given a vector of birthdays. We want to convert these into ages, using today's date as the reference point.

```{r}
birthday <- c("September 1, 1955", 
                 "06/21/1970",
                 "1992 July 17th",
                 "june 6 1980",
                 "1/4/1995",
                 "7/1/94",
                 "06/03/1992",
                 "7/28/00",
                 "1990/09/22",
                 "January 2, 1990",
                 "February22nd,  '98",
                 "June 3, 1982", 
                 "08-2-1999",
                 "2000-7-4")
```


#### Part (a)

There are issues with the data we are given - what are they?


#### Part (b)

Let's start cleaning this up - first let's decide the format that would be easiest to work with. 

The package `lubridate` can help us with this. I provided code below to calculate age using my birthday.

```{r}
refDate = Sys.Date()
birthDate = "1994-06-27"
time_length(difftime(refDate, birthDate), "years")
```

Here I used `YYYY-MM-DD` for my birthday, but there are other options in `lubridate`. We will stick with `YYYY-MM-DD` to simplify things. Let's start with the "easier" cases and go from there.

The first step is to find all dates of the form `DD/MM/YYYY`, `DD/MM/YY`, `DD-MM-YYYY`, etc. and create a new `data.frame` with 4 columns: `id`, `month`, `day`, and `year`. We'll do this together:

```{r}
ages = data.frame(id=1:length(birthday),dob=birthday)
pattern1 <- "^(([1-9])|(0[1-9])|(1[0-2]))(/|-)\\d{1,2}(/|-)\\d{2,4}$"
ages1 = ages %>% filter(str_detect(dob, pattern1)) %>%
  separate(dob, c("month", "day", "year"), sep = "/|-") %>%
  mutate_at(c("day","month"),str_remove, pattern="^0+") 
ages_left = ages %>% filter(!str_detect(dob, pattern1)) 
ages_left
```

Let's decode `pattern1`:

- `^` = start of string
- `|` = "or" operand 
- `()` captures groupings, i.e. `A(nt)|(pple)` will match Ant or Apple
- `(([1-9])|(0[1-9])|(1[0-2]))` = 1-9 or 01-09 or 10-12 (for month)
- `(/|-)` = / or -
- `\\d{1,2}` = one or two digits digit (for day)
- `\\d{2,4}` = two or four digits (for year)
- `$` = end of string

Next find the dates in numeric format with year first, and add those to `ages1`.


#### Part (c)

Now let's deal with the remaining cases. In all of them, month is spelled out (some with first letter capitalized, some without). In every entry, date immediately follows month. Year is sometimes first and sometimes last, with either 2 or 4 digits. Since day can only take values in 1-31, and two-digit years only take values in 50-98 (see `ages_left`) we can easily differentiate between 2 digit days and 2 digit years.

Try this out using the function `str_extract` and three different `patterns` for month, day, and year. You should be able to catch all the remaining entries.

#### Part (d)

Our parsed data is almost ready! We just have to fix a few last issues. For this problem, please fix the following:

1. The years that are 2 digits need to be 4 digits
2. All months need to be in numeric format
3. All single digits need to have a "0" in front

*Hint*: for each of `year`,`month`,`day` create a parse function which takes as input one string and converts it to the proper format. Then use `sapply` to apply it over every element in the `data.frame`. For example, `sapply(year, year_parse)`.

Next, add a new entry to your `data.frame` called dob. This should have the individual's birthday in the `YYYY-MM-DD` format so we can easily call the `compute_age` function below. You can do this using `paste(year,"-", month, "-", day, sep="")`.

```{r}
compute_age <- function(dob) {
  refDate = Sys.Date()
  return(time_length(difftime(refDate, dob), "years"))
}
```

### Problem 2: Weight

For this problem, you are given a vector of user-generated weights in pounds and kilograms. Convert this vector into a `data.frame` with three columns: `weight_original`, `weight_kg`, `weight_lbs`. Recall that 1 kg = 2.2046 lbs.

```{r}
weight <- c("55 kg", 
                "130 lbs",
                "150 pounds",
                "77 kilogram",
                "84kg", 
                "150 lb",
                "124 pound",
                "60 kg", 
                "70 kg")
  
```

### Problem 3: Extracting a table from a pdf

For this problem, we'll be using the table shown on the first page of <https://www.who.int/docs/default-source/coronaviruse/situation-reports/20200809-covid-19-sitrep-202.pdf?sfvrsn=2c7459f6_2>. It has a list of coronavirus cases and deaths by geographic region of the world. The first step is to import the table:

```{r import_table}
temp_file <- tempfile()
url <- "https://www.who.int/docs/default-source/coronaviruse/situation-reports/20200809-covid-19-sitrep-202.pdf?sfvrsn=2c7459f6_2"
download.file(url, temp_file)
txt <- pdf_text(temp_file)
file.remove(temp_file)
raw_covid <- txt[1]
raw_covid
```

Here we have an array of length 1, with the table contents written as a string. There's a bunch of text at the beginning that we don't need - so let's get rid of everything after "(new cases in last 24 hours)" using the `str_sub` and `str_locate` functions.

```{r split_string}
raw_covid <- raw_covid %>% str_sub(
  str_locate(raw_covid, "\\(new cases in last 24 hours\\)\n")[1,2]+1, nchar(raw_covid)
)
tab <- str_split(raw_covid, "\n")
tab <- tab[[1]]
tab
```

Now we have an array with 8 entries. The last entry is empty, but the others have a geographic region, followed by a bunch of space, followed by total overall cases, with new daily cases in parentheses, followed by a bunch of space, followed by total overall deaths, with new daily deaths in parentheses. There is also weird spacing in the numbers that we will want to fix. We want to transform this into a `data.frame` with 5 columns - `area`, `total_cases`, `daily_cases` `total_deaths`, `daily_deaths`.

#### Part (a) 
Create the `data.frame` described above. This involves splitting the string based on spaces of length > 2 using the `str_split` function, converting to a `data.frame`, naming the original three columns, removing the last empty row, separating `total_cases` into `total_cases` and `daily_cases` (hint: use the `separate` function with `sep="cases"`), and separating `total_deaths` into `total_deaths` and `daily_deaths` (hint: use the `separate` function with `sep="deaths"`). You can leave the numbers messy - we will fix this in the next question.

#### Part (b) 
Now that we have our `data.frame` we want to do two things: remove the spaces in the `total_cases`, `daily_cases`, `total_deaths`, and `daily_deaths` columns, and remove the parentheses in the `daily_cases` and `daily_deaths` columns (this can be done using `mutate_at` and `str_replace_all`) Then convert the numbers to numeric using `as.numeric` so we can manipulate them later. Recall that to match `(` and `)` you'll need to use `\\(` and `\\)` respectively, as these are special characters.


